Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2020-04-10 treestats_paper.tex
for resubmission

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We use this correspondence to define a general framework that
efficiently computes single-site population genetic statistics
using the succinct tree sequence encoding of genealogies and genome sequence.
The general approach accumulates ``sample weights'' within each marginal tree,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in two dual ways:
by \emph{site}, which corresponds to statistics calculated from genome
sequence; and by \emph{branch}, which gives the expected value of the site statistic
under the infinite-sites model of mutation.
We use the framework to implement many currently-defined statistics of genome
sequence (making the statistics' relationship to the underlying genealogical trees
concrete and explicit),
as well as the analogous ``branch'' statistics of tree shape. \revpoint{2}{10}
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
optimized matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may encode interesting biological
signals.

@@

Genotype matrices are massively redundant, however, and several specialized
compression methods have been proposed~\citep{
christley2008human,qiao2012handling,durbin2014efficient,sambo2014compression,layer2016efficient,
danek2018gtc,lin2019sparse}.
The fundamental reason for this redundancy is that samples share ancestry:
related individuals will tend to share the same state at a particular variant site.
Trees describe these ancestral
relationships~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encoding genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{hudson1983properties,griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share much of their structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, and has formed the basis for
efficient algorithms to calculate several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

The basic relationship between tree shape and summaries of genetic variation
that we will explain and utilize below is this:
if mutations are neutral, then the expected number that occur on a
given branch of a tree is proportional to the length of that branch.
This ``duality'' (we will use the term more precisely later) has been
used extensively in deriving properties of statistics in models of random mating
\citep[e.g.,][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
as has the fact that it applies regardless of the underlying demographic model
\citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
mcvean2002genealogical,lohse2016efficient,ralph2019empirical}. We build on this central
insight to define a flexible and computationally efficient framework for
population genetic statistics on tree sequences.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each marginal tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.

@@

Population genetic data is most directly represented as a matrix of genotypes,
and so summaries of population genetic data usually begin with this data structure.
However, since the genetic variation represented in a genotype matrix
was generated by mutations inherited through genealogical trees,
it is often informative to think about these statistics in terms of these trees.
Most population genetic summary statistics can be thought of in terms of the \emph{allele frequencies},
i.e., what proportion of the sampled genotypes carry each allele.
The frequency of any particular allele
can easily be found from the location of the mutation(s) that produced it
on the marginal genealogical tree,
by counting the number of sampled genotypes below it in the tree.
If many mutations have occurred on a single genealogical tree,
then these mutations start to give us an idea of the shape of that tree.
This relationship is depicted in \autoref{fig:allele_freqs},
which shows the relationship between the genealogical tree
and the frequencies of alleles at five different sites across a region of the genome
described by that single tree.
Our goal will be to generalize this sort of summary of the genotype matrix,
using the underlying genealogical trees.
Working with the trees serves two purposes.
First, the trees are closer to the demographic processes that we are really interested in,
and so allow us to reason more directly about population genetic signal and noise.
Second, trees are highly efficient data structures,
and allow us to compute with genetic data far more efficiently than with the genotype matrix itself.

\begin{figure}
    \begin{center}
        \includegraphics{figures/allele_frequency_diagram}
    \end{center}
    \caption{
        The relationship between allele frequencies and the underlying genealogical trees.
        \textbf{(a)} The genotype matrix, showing genotypes of five sequences at five separate sites,
        as well as the derived allele frequencies (derived alleles are shown in color,
        with separate colors for each genotype pattern).
        \textbf{(b)} The genealogical tree describing how those five samples are related to each other
        over this segment of genome. The samples are leaves,
        and each mutation is represented as a star on the genealogy,
        with color matching the genotype matrix, and
        labeled with its index in the genotype matrix
        and the ancestral and derived states at that site.
        \textbf{(c)} Each node in the tree is labeled with the number of samples
        at or below that node, which we think of as a \emph{weight}.
        Note that this is \emph{additive}:
        the weight of each node is the sum of the weights of its child nodes.
        Since the frequency of a derived allele is the number of samples
        that have inherited that mutation, the allele frequency of each mutation
        is equal to the weight of the node directly below the mutation,
        divided by the total sample size.
        \label{fig:allele_freqs}
    }
\end{figure}

The operation of finding allele frequencies on a tree,
depicted in \autoref{fig:allele_freqs}(c)
is the basis of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
Somewhat more generally, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch.
Suppose we want to calculate one of the linear functions of the frequency spectrum
discussed by \citet{fu1995statistical} and \citet{achaz2009frequency},
which can be described as follows.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple sum over sites,
we can compute this by summing over the $L$ sites in the genome and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
For instance, mean genetic diversity across biallelic loci is computed using this formula
with $f(p) = p (1-p)$,
omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
This is because genetic diversity is defined to be the probability that two randomly chosen sequences
differ at a randomly chosen site,
and the probability that two randomly chosen genomes differ at a site with allele frequency $p$
is $p (1-p) + (1-p) p$.
If we know where each mutation has occurred on each marginal genealogical tree,
we can easily compute statistics of this form
by finding allele frequencies, applying the function $f( )$ to each,
and summing up the results.

@@

\begin{definition}[Sample and subtree weights] \label{defn:weights}
    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
    to every sample node.
    Given these weights, on tree $T_k$
    the \emph{subtree weight} $\nw_k(u)$
    for a node $u$ is the sum of all sample weights
    of every sample node that is descended from $u$ in the tree (including $u$, if it is a sample):
    \begin{align*}
        \nw_k(u) = \sum_{v \,:\, v \le_{T_k} u} \iw(v) ,
    \end{align*}
    where $v \le_{T_k} u$ if $u$ is on the path from $v$ to root in the tree $T_k$.
    The \emph{total weight} is the sum of the weights over all samples:
    $\tiw = \sum_v \iw(v)$.
\end{definition}

@@

Perhaps the most natural way to summarize weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us, for each node, what proportion of the sample's genomes were inherited from that node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
and to the remaining weight of the rest of the tree,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{|\treeseq|} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the sum of the weights of all samples descending from node $u$
in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all remaining weights.
If this is computed over only a ``window'' of the genome
then $L_k$ is replaced by the length of the portion of that window that the tree $T_k$ extends for,
and $L$ is replaced by the length of the window.

@@

\begin{example}[Ancestry proportions] \label{ex:ancestry_props} \revpoint{2}{5}
    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
    then $\nw(u) / n$ is the proportion of the samples in $S$ that inherit from $u$.
    Therefore, if $f(x) = x / n$,
    then $\nodep(f, \iw)_u$ is the proportion of the genomes of $S$
    that are inherited from (ancestor) $u$.
\end{example}

@@

If there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
In the more general case of recurrent and back mutations, $\aw$ can
be computed from subtree weights in a straightforward way.
We then define the \textbf{site statistic} of site $j$ for a summary function $f()$
and sample weights $\iw$ to be the sum of $f()$
applied to the weight of every allele, $a$, found at this site:
\begin{align}
    \site(f, \iw)_j
    &=
    \sum_{a} f(\aw_j(a)).
\end{align}
Here, we think of a ``site'' as a single nucleotide position on the genome.
We often want to summarize statistics across regions of the genome (``windows'').
To do this, we overload notation somewhat and use a subscript $[i,j)$ to denote an average
over the corresponding portion of the genome:
\begin{align}
    \site(f, \iw)_{[i,j)}
    &=
    \frac{1}{j-i} \sum_{k=i}^{j-1} \site(f, \iw)_k
\end{align}
The requirement on strict summary functions that $f(0) = f(\tiw) = 0$
ensures that the sum is only affected by polymorphic sites,
although we normalize by total number of sites, to allow comparison between regions.

@@

\begin{example}[Nucleotide diversity] \label{ex:site_diversity}
    The nucleotide diversity of a group $S$ of $n$ samples
    is the average density of differences between pairs of samples,
    or equivalently, the average heterozygosity across positions.
    To calculate this statistic,
    % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
    let $\iw = \bone_S$,
    so that $\nw(u)$ gives the number of nodes in $S$ inheriting from $u$,
    and define
    \begin{align*}
        f(x) = \frac{x (n - x)}{n (n-1)} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity of $S$ in the region between $a$ and $b$.
\end{example}

@@

\begin{example}[Nucleotide divergence] \label{ex:site_divergence}
    Now suppose we want to compute the mean pairwise density of nucleotide differences
    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
    with $n_1$ and $n_2$ samples, respectively.
    As before,
    let $\iw_j = \bone_{S_j}$,
    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$,
    and define
    \begin{align*}
        f(x_1, x_2) = \frac{x_1}{n_1} \left(1 - \frac{x_2}{n_2}\right) .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
    in the region between $a$ and $b$.
\end{example}

@@

\begin{figure}
    \begin{center}
    \includegraphics{figures/divergence_diagram_simple}
    \end{center}
    \caption{
        An example of computing sequence divergence
        between red samples (triangles) and blue samples (squares)
        at a single site with a T$\to$G mutation, using the tree.
        The derived mutation is found in $a=2$ out of the $n_\text{red} = 3$ red samples
        and in $b=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
        but a randomly chosen blue node does not
        is given by the summary function, $f(a,b) = f(2, 1) = 2/6$.
        The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
        a randomly chosen red and blue pair, but with the blue node carrying the derived allele,
        so the total probability that a randomly chosen pair
        differs at this site is $f(2,1) + f(1,1) = 1/2$.
        In this example, subtree weights $(a,b)$ record how many red samples ($a$)
        and how many blue samples ($b$) lie in that subtree,
        and the weights at each node are equal to the sum of the weights of the node's children,
        as in \autoref{fig:adding_weights}.
        \label{fig:divergence_diagram}
    }
\end{figure}

\begin{example}[Segregating sites] \label{ex:segregating_sites}
    Again, let $\iw = \bone_{S}$ for a group of $n$ samples,
    and now define
    \begin{align*}
        f(x) = \begin{cases}
            1 - \frac{x}{n} \qquad &\text{if } x > 0 \\
            0 \qquad &\text{otherwise} .
        \end{cases}
    \end{align*}
    In computing $\site(f, \iw)_{[a,b)}$, the function $f()$ is given the number of samples in $S$
    that inherit each allele.
    A site with $k$ distinct alleles that segregate in $S$ with counts $x_1 + \cdots + x_k = n$
    contributes $f(x_1) + \cdots + f(x_k) = (1 - x_1/n) + \cdots + (1 - x_k/n) = k - 1$
    to the statistic. \revpoint{2}{6}
    Therefore, $\site(f, \iw)_{[a,b)}$, is the minimum number of derived mutations per unit length,
    which is the density of segregating sites if all sites are biallelic.
    (The actual number of mutations per unit length will be greater if there have been back mutations.)
\end{example}

\paragraph{Polyallelic sites}
The summary function in Example~\ref{ex:segregating_sites} might be surprising:
the reason for its particular form is so that the statistic returns a sensible answer
even when sites may have more than two alleles.
There is not a consensus in the field
on how to use sites with more than two alleles, but the most common practice is
to reduce to biallelic data, by either discarding other sites
or by marking all non-ancestral alleles as (the same) ``derived'' allele.
We choose to define statistics in a way that still makes formal sense for polyallelic sites,
so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
This is not the only choice, but is natural because it agrees with what is obtained by
looking at pairwise haplotype differences.
For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
gives the probability that a randomly chosen sample from each set differ,
a definition that makes sense even with polyallelic sites.

@@

This relationship is true more generally.
In fact, for any region of the genome between $i$ and $j$,
% with $\treeseq_{[i,j)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[i,j)}
    =
    \frac{1}{\mu}
    \E\left[ \site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)} \right].
\end{align}
Here, $\treeseq_{[i,j)}$ denotes the tree sequence on the interval $[i,j)$,
and so the expectation keeps the genealogies fixed
and averages over infinite-sites mutations
with rate $\mu$ per unit time and per unit of sequence length.
Note that the expected product of two site statistics,
$\E[\site(f, \iw) \site(g, \iw')]$
is not equal to the product of the two branch statistics, $\branch(f, \iw) \branch(g, \iw')$,
because they are not independent.
However, it is always possible to define a branch statistic that
gives the expected value of the product~\citep{ralph2019empirical}.

>>>>

As a genetic mutation is passed down across generations,
it distinguishes those genomes that have inherited it from those that have not,
providing a glimpse of the genealogical tree relating the genomes to each other at that site.
Statistical summaries of genetic variation therefore also describe the underlying genealogies.
We use this correspondence to define a general framework that
efficiently computes single-site population genetic statistics
using the succinct tree sequence encoding of genealogies and genome sequence.
The general approach accumulates ``sample weights''
within the genealogical tree at each position on the genome,
which are then combined using a ``summary function'';
different statistics result from different choices of weight and function.
Results can be reported in three ways:
by \emph{site}, which corresponds to statistics calculated as usual from genome sequence;
by \emph{branch}, which gives the expected value of the dual site statistic
under the infinite-sites model of mutation,
and by \emph{node}, which summarizes the contribution of each ancestor to these statistics.
We use the framework to implement many currently-defined statistics of genome sequence
(making the statistics' relationship to the underlying genealogical trees concrete and explicit),
as well as the corresponding ``branch'' statistics of tree shape. \revpoint{2}{10}
We evaluate computational performance using simulated data, and
show that calculating statistics from tree sequences using this general
framework is several orders of magnitude more efficient than
optimized matrix-based methods in terms of both run time and memory requirements.
We also explore how well the duality between site and branch statistics holds
in practice on trees inferred from the 1000 Genomes Project dataset,
and discuss ways in which deviations may encode interesting biological signals.

@@

Genotype matrices are massively redundant, however, and several specialized
compression methods have been proposed~\citep{
christley2008human,qiao2012handling,durbin2014efficient,sambo2014compression,layer2016efficient,
danek2018gtc,lin2019sparse}.
The fundamental reason for this redundancy is that samples share ancestry:
related individuals will tend to share the same state at a particular variant site.
Trees describe these ancestral
relationships~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encoding genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually reproducing organisms, because
recombination results in many trees along the
genome~\citep{hudson1983properties,griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share much of their structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, and formed the basis for
efficient algorithms to calculate several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach of \citet{kelleher2016efficient} to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation,
including polyallelic sites and back mutations.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

The basic relationship between tree shape and summaries of genetic variation
that we will explain and utilize below is this:
if mutations are neutral, then the expected number that occur on a
given branch of a tree is proportional to the length of that branch.
This ``duality'' (we will use the term more precisely later) has been
used extensively in deriving properties of statistics in models of random mating
\citep[e.g.,][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
as has the fact that it applies regardless of the underlying demographic model
\citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
mcvean2002genealogical,lohse2016efficient,ralph2019empirical}. We build on this central
insight to define a flexible and computationally efficient framework for
population genetic statistics on tree sequences.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each marginal tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
Different choices of weights and summary functions
will then produce both familiar and novel statistics
of both genotypes and genealogies.

@@

Population genetic data is most directly represented as a matrix of genotypes,
and so summaries of population genetic data usually begin with this data structure.
However, since the genetic variation represented in a genotype matrix
was generated by mutations inherited through genealogical trees,
it is often informative to think about these statistics in terms of these trees.
Most population genetic summary statistics can be thought of in terms of the \emph{allele frequencies},
i.e., what proportion of some subset of the sampled genotypes carry each allele.
The frequency of any particular allele
can easily be found from the location of the mutation(s) that produced it
on the marginal genealogical tree,
by counting the number of sampled genotypes below the mutation(s) in the tree.
If many mutations have occurred at distinct sites that share a single genealogical tree,
then these mutations start to give us an idea of the shape of that tree.
This relationship is depicted in \autoref{fig:allele_freqs},
which shows the relationship between the genealogical tree
and the frequencies of alleles at five different sites across a region of the genome
described by that single tree.
Our goal will be to generalize this sort of summary of the genotype matrix,
using the underlying genealogical trees.
Working with the trees serves two purposes.
First, the trees are closer to the demographic processes that we are really interested in,
and so allow us to reason more directly about population genetic signal and noise.
Second, trees are highly efficient data structures,
and allow us to compute with genetic data far more efficiently than with the genotype matrix itself.

\begin{figure}
    \begin{center}
        \includegraphics{figures/allele_frequency_diagram}
    \end{center}
    \caption{
        The relationship between allele frequencies and the underlying genealogical trees.
        \textbf{(a)} The genotype matrix, showing genotypes of five sequences (``samples'')
        at five separate sites,
        as well as the derived allele frequencies (derived alleles are shown in color,
        with separate colors for each genotype pattern).
        \textbf{(b)} The genealogical tree describing how those five samples are related to each other
        over this segment of genome. The samples are leaves,
        and each mutation is represented as a star on the genealogy,
        with color matching the genotype matrix, and
        labeled with its index in the genotype matrix
        and the ancestral and derived states at that site.
        \textbf{(c)} Each node in the tree is labeled with the number of samples
        at or below that node, which we think of as a \emph{weight}.
        Note that this is \emph{additive}:
        the weight of each node is the sum of the weights of its child nodes.
        Since the frequency of a derived allele is the number of samples
        that have inherited that mutation, the allele frequency of each mutation
        is equal to the weight of the node directly below the mutation,
        divided by the total sample size.
        \label{fig:allele_freqs}
    }
\end{figure}

The operation of finding allele frequencies on a tree,
depicted in \autoref{fig:allele_freqs}(c)
is the basis of a familiar operation in population genetics:
summarizing the allele frequency spectrum.
Somewhat more generally, we could calculate how many of a certain set $A$ of samples
inherit from a particular branch in a tree
by (a) assigning each of these samples weight 1 (and other samples weight zero), then
(b) finding the total weight of all samples in the subtree inheriting from that branch.
This gives the number of samples that would inherit any mutation falling on this branch.
Suppose we want to calculate one of the linear functions of the frequency spectrum
discussed by \citet{fu1995statistical} and \citet{achaz2009frequency},
which can be described as follows.
Write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
--- i.e., the number of alleles whose frequency is $p$ in our sample ---
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
Since this is a simple sum over sites,
we can compute this by summing over the $L$ sites in the genome and all alleles
(including the ancestral allele),
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
For instance, mean genetic diversity across biallelic loci is computed using this formula
with $f(p) = p (1-p)$,
omitting the usual factor of two because the sum is over \emph{both} alleles at each site.
This is because genetic diversity is defined to be the probability that two randomly chosen sequences
differ at a randomly chosen site,
and the probability that two randomly chosen genomes differ at a site with allele frequency $p$
is $p (1-p) + (1-p) p$.
If we know where each mutation has occurred on the genealogical tree at each site in the genome,
we can easily compute statistics of this form
by finding allele frequencies, applying the function $f( )$ to each,
and summing up the results.

@@

\begin{definition}[Sample and subtree weights] \label{defn:weights}
    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
    to every sample node.
    Given these weights,
    the \emph{subtree weight} $\nw_k(u)$ on tree $T_k$
    for a node $u$ is the sum of all sample weights
    of every sample node that is descended from $u$ in the tree (including $u$, if it is a sample):
    \begin{align*}
        \nw_k(u) = \sum_{v \,:\, v \le_{T_k} u} \iw(v) ,
    \end{align*}
    where $v \le_{T_k} u$ if $u$ is on the path from $v$ to root in the tree $T_k$.
    The \emph{total weight} is the sum of the weights over all samples:
    $\tiw = \sum_v \iw(v)$.
\end{definition}

@@

Perhaps the most natural way to summarize weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us, for each node, what proportion of the sample's genomes were inherited from that node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
and to the remaining weight of the rest of the tree,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{|\treeseq|} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the sum of the weights of all samples descending from node $u$
in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all remaining weights.
This is an average over the genome because each tree is weighted by its span, $L_k$,
and the sum is divided by the total genome length,
which is the sum of all the spans: $L = L_1 + \cdots + L_{|\treeseq|}$.
If the node statistic is computed over only a ``window'' of the genome
then $L_k$ is replaced by the length of the portion of that window that the tree $T_k$ extends for,
and $L$ is replaced by the length of the window.

@@

\begin{example}[Ancestry proportions] \label{ex:ancestry_props} \revpoint{2}{5}
    If $\iw = \bone_S$ are the indicator weights of the set $S$ of $n$ samples,
    then $\nw(u) / n$ is the proportion of the samples in $S$ that inherit from $u$.
    Therefore, if $f(x) = x / n$,
    then $\nodep(f, \iw)_u$ is the proportion of the genomes of $S$
    that are inherited from (ancestor) $u$.
\end{example}

@@

If there has been only one mutation at the site,
then $\aw_j(a)$ is equal to the weight of the subtree inheriting from the mutation that produced $a$.
In the more general case of recurrent and back mutations, $\aw$ can
be computed from subtree weights in a straightforward way.
We then define the \textbf{site statistic} of site $j$ for a summary function $f()$
and sample weights $\iw$ to be the sum of $f()$
applied to the weight of every allele, $a$, found at this site:
\begin{align}
    \site(f, \iw)_j
    &=
    \sum_{a} f(\aw_j(a)).
\end{align}
Here, we think of a ``site'' as a single nucleotide position on the genome.
We often want to summarize statistics across regions of the genome (``windows'').
To do this, we overload notation somewhat and use a subscript $[i,j)$ to denote an average
over the corresponding portion of the genome:
\begin{align}
    \site(f, \iw)_{[i,j)}
    &=
    \frac{1}{j-i} \sum_{k=i}^{j-1} \site(f, \iw)_k
\end{align}
The requirement on strict summary functions that $f(0) = f(\tiw) = 0$
ensures that the sum is only affected by polymorphic sites,
although we normalize by total number of sites,
so that the values are comparable between different regions of the genome.

@@

\begin{example}[Nucleotide diversity] \label{ex:site_diversity}
    The nucleotide diversity of a group $S$ of $n$ samples
    is the average density of differences between pairs of samples,
    or equivalently, the average heterozygosity across positions.
    To calculate this statistic,
    % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
    let $\iw = \bone_S$,
    so that $\nw(u)$ gives the number of nodes in $S$ inheriting from $u$,
    and define
    \begin{align*}
        f(x) = \frac{x (n - x)}{n (n-1)} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity of $S$
    in the region of the genome between $a$ and $b$.
\end{example}

@@

\begin{example}[Nucleotide divergence] \label{ex:site_divergence}
    Now suppose we want to compute the mean pairwise density of nucleotide differences
    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
    with $n_1$ and $n_2$ samples, respectively.
    As before,
    let $\iw_j = \bone_{S_j}$,
    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$,
    and define
    \begin{align*}
        f(x_1, x_2) = \frac{x_1}{n_1} \left(1 - \frac{x_2}{n_2}\right) .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
    in the region of the genome between $a$ and $b$.
\end{example}

@@

\begin{figure}
    \begin{center}
    \includegraphics{figures/divergence_diagram_simple}
    \end{center}
    \caption{
        An example of computing sequence divergence
        between red samples (triangles) and blue samples (squares)
        at a single site with a T$\to$G mutation, using the tree.
        Each node is labeled with subtree weights $(x_\text{red},x_\text{blue})$,
        recording how many red samples ($x_\text{red}$)
        and how many blue samples ($x_\text{blue}$) lie in that subtree.
        The derived mutation is found in $x_\text{red}=2$ out of the $n_\text{red} = 3$ red samples
        and in $x_\text{blue}=1$ of the $n_\text{blue}=2$ blue samples, and so the probability that a randomly chosen red node carries the derived allele
        but a randomly chosen blue node does not
        is given by the summary function, $f(x_\text{red}, x_\text{blue}) = f(2, 1) = 2/6$.
        The complementary value, $f(1, 1) = 1/6$ is the probability that the mutation separates
        a randomly chosen red and blue pair, but with the blue node carrying the derived allele,
        so the total probability that a randomly chosen pair
        differs at this site is $f(2,1) + f(1,1) = 1/2$.
        \label{fig:divergence_diagram}
    }
\end{figure}

\begin{example}[Segregating sites] \label{ex:segregating_sites}
    Again, let $\iw = \bone_{S}$ for a group of $n$ samples,
    and now define
    \begin{align*}
        f(x) = \begin{cases}
            1 - \frac{x}{n} \qquad &\text{if } x > 0 \\
            0 \qquad &\text{otherwise} .
        \end{cases}
    \end{align*}
    In computing $\site(f, \iw)_{[a,b)}$, the argument $x$ to the function $f(x)$
    is the number of samples in $S$ that inherit each allele.
    A site with $k$ distinct alleles that segregate in $S$ with counts $x_1 + \cdots + x_k = n$
    contributes $f(x_1) + \cdots + f(x_k) = (1 - x_1/n) + \cdots + (1 - x_k/n) = k - 1$
    to the statistic. \revpoint{2}{6}
    Therefore, $\site(f, \iw)_{[a,b)}$, is the minimum number of derived mutations per unit length,
    which is the density of segregating sites if all sites are biallelic.
    (The actual number of mutations per unit length will be greater if there have been back mutations.)
\end{example}

\paragraph{Polyallelic sites}
The summary function in Example~\ref{ex:segregating_sites} might be surprising:
the reason for its particular form is so that the statistic returns a sensible answer
even when sites may have more than two alleles.
There is not a consensus in the field
on how to use sites with more than two alleles, but the most common practice is
to reduce to biallelic data, by either discarding other sites
or by marking all non-ancestral alleles as (the same) ``derived'' allele.
In contrast, we choose to define statistics in a way that still makes formal sense for polyallelic sites,
so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
This is not the only choice, but is natural because it agrees with what is obtained by
looking at pairwise haplotype differences.
For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
gives the probability that a randomly chosen sample from each set differ,
a definition that makes sense even with polyallelic sites.

@@

This relationship is true more generally.
In fact, for any region of the genome between $i$ and $j$,
% with $\treeseq_{[i,j)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[i,j)}
    =
    \frac{1}{\mu}
    \E\left[ \site(f, \iw)_{[i,j)} \given \treeseq_{[i,j)} \right].
\end{align}
Here, $\treeseq_{[i,j)}$ denotes the tree sequence on the interval $[i,j)$,
and so the expectation keeps the genealogies fixed
and averages over infinite-sites mutations
with rate $\mu$ per unit time and per unit of sequence length.
Note that the expected product of two site statistics,
$\E[\site(f, \iw) \site(g, \iw') \given \treeseq]$
is not equal to the product of the two branch statistics, $\mu^2 \branch(f, \iw) \branch(g, \iw')$,
because they are not independent.
However, it is always possible to define a branch statistic that
gives the expected value of the product, as described in \citet{ralph2019empirical}.
